---
title: "MAST30027 Modern Applied Statistics Assignment 3"
author: "James La Fontaine"
subtitle: "Tutorial:Wed 1-2PM, Yidi Deng"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;

\pagebreak

# Question 1d

```{r load in data}
X = scan(file="assignment3_prob1_2023.txt", what=double())

length(X)

hist(X)
```

# Implementation of the EM algorithm
We will assume that the observed data follows a mixture of three Poisson distributions. Specifically, for $i=1 \ldots, n$,
\[
Z_i \sim \mbox{categorical }(\pi_1, \pi_2, 1-\pi_1-\pi_2),
\]
\[
 \begin{aligned}
  X_i | Z_i = 1 &\sim \mbox{Poisson}(\lambda_1), \\
  X_i | Z_i = 2 &\sim \mbox{Poisson}(\lambda_2), \\
  X_i | Z_i = 3 &\sim \mbox{Poisson}(\lambda_3).
  \end{aligned}
\]
We aim to obtain MLE of parameters $\theta = (\pi_1, \pi_2, \lambda_1, \lambda_2, \lambda_3)$ using the EM algorithm.

We implement the E and M step in the `EM.iter` function below. The `compute.log.lik` function below computes the incomplete log-likelihood, assuming the parameters are known. We will check that the incomplete log-likelihoods increases at each step by plotting them. The `mixture.EM` function is the main function which runs multiple EM steps and checks for convergence by computing the incomplete log-likelihoods at each step.
```{r}
# w.init : initial value for pi
# lambda.init : initial value for lambda
# epsilon : If the incomplete log-likelihood has changed by less than epsilon, 
# EM will stop.
# max.iter : maximum number of EM-iterations
mixture.EM <- function(X, w.init, lambda.init, epsilon=1e-5, max.iter=100) {
  
  w.curr = w.init
  lambda.curr = lambda.init
  
  # store incomplete log-likehoods for each iteration
  log_liks = c()
  
  # compute incomplete log-likehoods using initial values of parameters. 
  log_liks = c(log_liks, compute.log.lik(X, w.curr, lambda.curr)$ill)
  
  # set the change in incomplete log-likelihood with 1
  delta.ll = 1
  
  # number of iteration
  n.iter = 1
  
  # If the log-likelihood has changed by less than epsilon, EM will stop.   
  while((delta.ll > epsilon) & (n.iter <= max.iter)){
    
    # run EM step
    EM.out = EM.iter(X, w.curr, lambda.curr)
    
    # replace the current value with the new parameter estimate
    w.curr = EM.out$w.new
    lambda.curr = EM.out$lambda.new
    
    # incomplete log-likehoods with new parameter estimate 
    log_liks = c(log_liks, compute.log.lik(X, w.curr, lambda.curr)$ill)
    
    # compute the change in incomplete log-likelihood 
    delta.ll = log_liks[length(log_liks)]  - log_liks[length(log_liks)-1]
    
    # increase the number of iteration 
    n.iter = n.iter + 1
  }
  return(list(w.curr=w.curr, lambda.curr=lambda.curr, log_liks=log_liks))
}
```

```{r}
EM.iter <- function(X, w.curr, lambda.curr) {
  
  # E-step: compute E_{Z|X,\theta_0}[I(Z_i = k)]
  
  # for each sample $X_i$, compute $P(X_i, Z_i=k)$ 
  prob.x.z = compute.prob.x.z(X, w.curr, lambda.curr)$prob.x.z
  
  # compute P(Z_i=k | X_i)
  P_ik = prob.x.z / rowSums(prob.x.z)
  
  # M-step
  w.new = colSums(P_ik)/sum(P_ik)  # sum(P_ik) is equivalent to sample size 
  lambda.new = colSums(P_ik*X)/colSums(P_ik)
  
  return(list(w.new=w.new, lambda.new=lambda.new))
}

```

Now we write a function to compute the incomplete log-likelihood, assuming the parameters are known.
$$\ell(\theta) = \sum_{i=1}^n \log \left( \sum_{k=1}^3 \pi_k f(x_i;\lambda_k) \right )$$

```{r}
# Compute incomplete log-likehoods
compute.log.lik <- function(X, w.curr, lambda.curr) {
  
  # for each sample $X_i$, compute $P(X_i, Z_i=k)$
  prob.x.z = compute.prob.x.z(X, w.curr, lambda.curr)$prob.x.z
  
  # incomplete log-likehoods
  ill = sum(log(rowSums(prob.x.z)))
  
  return(list(ill=ill))
}

# for each sample $X_i$, compute $P(X_i, Z_i=k)$
compute.prob.x.z <- function(X, w.curr, lambda.curr) {
  
  # for each sample $X_i$, compute $P(X_i, Z_i=k)$. Store these values in the columns of L:
  L = matrix(NA, nrow=length(X), ncol= length(w.curr))
  for(k in seq_len(ncol(L))) {
    L[, k] = dpois(X, lambda=lambda.curr[k])*w.curr[k]
  }
  
  return(list(prob.x.z=L))
}
```

# Apply the EM algorithm

Run EM algorithm with different initial values and check that the incomplete log-likelihoods increases at each step by plotting them. 
```{r, fig.width = 4, fig.height = 4}
EM1 <- mixture.EM(X, w.init=c(0.3,0.3, 0.4), lambda.init=c(3, 20, 35), epsilon=1e-5, max.iter=100)
ee = EM1
print(paste("Estimate pi = (", round(ee$w.curr[1],2), ",",
            round(ee$w.curr[2],2), ",",
            round(ee$w.curr[3],2), ")", sep=""))

print(paste("Estimate lambda = (", round(ee$lambda.curr[1],2), ",",
            round(ee$lambda.curr[2],2), ",",
            round(ee$lambda.curr[3],2), ")", sep=""))

plot(ee$log_liks, ylab='incomplete log-likelihood', xlab='iteration')
```

```{r, fig.width = 4, fig.height = 4}
EM2 <- mixture.EM(X, w.init=c(0.1,0.2, 0.7), lambda.init=c(5, 25, 40), epsilon=1e-5, max.iter=100)
ee = EM2
print(paste("Estimate pi = (", round(ee$w.curr[1],2), ",",
            round(ee$w.curr[2],2), ",",
            round(ee$w.curr[3],2), ")", sep=""))

print(paste("Estimate lambda = (", round(ee$lambda.curr[1],2), ",",
            round(ee$lambda.curr[2],2), ",",
            round(ee$lambda.curr[3],2), ")", sep=""))

plot(ee$log_liks, ylab='incomplete log-likelihood', xlab='iteration')
```

Check which estimators have the highest incomplete log-likelihood.
```{r}
EM1$log_liks[length(EM1$log_liks)]
EM2$log_liks[length(EM2$log_liks)]
```

Estimators from the two EM runs have (equally) highest incomplete log-likelihoods. You can see that the estimators from the EM runs are the same, so it doesn’t matter which estimators we choose. Lets choose the estimators from the first EM run - $\hat\pi_1 = 0.25,\; \hat\pi_2 = 0.25,\; \hat\lambda_1 = 5.17,\; \hat\lambda_2 = 18.09,\; \hat\lambda_3 = 36.94$.

\pagebreak

# Question 2c

```{r load in data 2}
X = scan(file="assignment3_prob1_2023.txt", what=double())

X0 = scan(file="assignment3_prob2_2023.txt", what=double())

length(X)

length(X0)

par(mfrow=c(1,2))

hist(X)

hist(c(X,X0))
```

# Implementation of the EM algorithm
We will assume that the observed data follows a mixture of three Poisson distributions. Additionally, for samples 301 to 400, we assume that they follow a known Poisson distribution. Specifically, for $i=1 \ldots, n$,
\[
Z_i \sim \mbox{categorical }(\pi_1, \pi_2, 1-\pi_1-\pi_2),
\]
\[
 \begin{aligned}
  X_i | Z_i = 1 &\sim \mbox{Poisson}(\lambda_1), \\
  X_i | Z_i = 2 &\sim \mbox{Poisson}(\lambda_2), \\
  X_i | Z_i = 3 &\sim \mbox{Poisson}(\lambda_3), \\ 
  \end{aligned}
\]
  and for i=301, ..., 400,
\[
X_i \sim \mbox{Poisson}(\lambda_2). \\
\]
We aim to obtain MLE of parameters $\theta = (\pi_1, \pi_2, \lambda_1, \lambda_2, \lambda_3)$ using the EM algorithm.

We implement the E and M step in the `EM.iter` function below. The `compute.log.lik` function below computes the incomplete log-likelihood, assuming the parameters are known. We will check that the incomplete log-likelihoods increases at each step by plotting them. The `mixture.EM` function is the main function which runs multiple EM steps and checks for convergence by computing the incomplete log-likelihoods at each step.
```{r}
# w.init : initial value for pi
# lambda.init : initial value for lambda
# epsilon : If the incomplete log-likelihood has changed by less than epsilon, 
# EM will stop.
# max.iter : maximum number of EM-iterations
mixture.EM <- function(X, w.init, lambda.init, epsilon=1e-5, max.iter=100) {
  
  w.curr = w.init
  lambda.curr = lambda.init
  
  # store incomplete log-likehoods for each iteration
  log_liks = c()
  
  # compute incomplete log-likehoods using initial values of parameters. 
  log_liks = c(log_liks, compute.log.lik(X, w.curr, lambda.curr)$ill)
  
  # set the change in incomplete log-likelihood with 1
  delta.ll = 1
  
  # number of iteration
  n.iter = 1
  
  # If the log-likelihood has changed by less than epsilon, EM will stop.   
  while((delta.ll > epsilon) & (n.iter <= max.iter)){
    
    # run EM step
    EM.out = EM.iter(X, w.curr, lambda.curr)
    
    # replace the current value with the new parameter estimate
    w.curr = EM.out$w.new
    lambda.curr = EM.out$lambda.new
    
    # incomplete log-likehoods with new parameter estimate 
    log_liks = c(log_liks, compute.log.lik(X, w.curr, lambda.curr)$ill)
    
    # compute the change in incomplete log-likelihood 
    delta.ll = log_liks[length(log_liks)]  - log_liks[length(log_liks)-1]
    
    # increase the number of iteration 
    n.iter = n.iter + 1
  }
  return(list(w.curr=w.curr, lambda.curr=lambda.curr, log_liks=log_liks))
}
```

```{r}
EM.iter <- function(X, w.curr, lambda.curr) {
  
  # E-step: compute E_{Z|X,\theta_0}[I(Z_i = k)]
  
  # for each sample $X_i$, compute $P(X_i, Z_i=k)$ 
  prob.x.z = compute.prob.x.z(X, w.curr, lambda.curr)$prob.x.z
  
  # compute P(Z_i=k | X_i)
  P_ik = prob.x.z / rowSums(prob.x.z)
  
  # M-step
  w.new = colSums(P_ik)/sum(P_ik)  # sum(P_ik) is equivalent to sample size 
  lambda.new = colSums(P_ik*X)/colSums(P_ik)
  lambda.new[2] = (colSums(P_ik*X)[2] + sum(X0))/(colSums(P_ik)[2] + 100)
  
  #print(paste("lambda.new = (", lambda.new, ")", sep=""))

  
  return(list(w.new=w.new, lambda.new=lambda.new))
}

```

Now we write a function to compute the incomplete log-likelihood, assuming the parameters are known.
$$\ell(\theta) = \sum_{i=1}^n \log \left( \sum_{k=1}^3 \pi_k f(x_i;\lambda_k) \right)$$
<!-- PUT THIS IN ABOVE DOLLAR SIGNS IF INC. LL DOES CHANGE FOR Q2 -> +  \sum_{i=301}^{400} \log(f(x_i;\lambda_2)) --> 

 

```{r}
# Compute incomplete log-likehoods
compute.log.lik <- function(X, w.curr, lambda.curr) {
  
  # for each sample $X_i$, compute $P(X_i, Z_i=k)$
  prob.x.z = compute.prob.x.z(X, w.curr, lambda.curr)$prob.x.z
  
  # incomplete log-likehoods
  ill = sum(log(rowSums(prob.x.z)))
  
  return(list(ill=ill))
}

# for each sample $X_i$, compute $P(X_i, Z_i=k)$
compute.prob.x.z <- function(X, w.curr, lambda.curr) {
  
  # for each sample $X_i$, compute $P(X_i, Z_i=k)$. Store these values in the columns of L:
  L = matrix(NA, nrow=length(X), ncol= length(w.curr))
  for(k in seq_len(ncol(L))) {
    L[, k] = dpois(X, lambda=lambda.curr[k])*w.curr[k]
  }
  
  return(list(prob.x.z=L))
}
```

# Apply the EM algorithm

Run EM algorithm with different initial values and check that the incomplete log-likelihoods increases at each step by plotting them. 
```{r, fig.width = 4, fig.height = 4}
EM1 <- mixture.EM(X, w.init=c(0.3,0.3, 0.4), lambda.init=c(3, 20, 35), epsilon=1e-5, max.iter=100)
ee = EM1
print(paste("Estimate pi = (", round(ee$w.curr[1],2), ",",
            round(ee$w.curr[2],2), ",",
            round(ee$w.curr[3],2), ")", sep=""))

print(paste("Estimate lambda = (", round(ee$lambda.curr[1],2), ",",
            round(ee$lambda.curr[2],2), ",",
            round(ee$lambda.curr[3],2), ")", sep=""))

plot(ee$log_liks, ylab='incomplete log-likelihood', xlab='iteration')
```

```{r, fig.width = 4, fig.height = 4}
EM2 <- mixture.EM(X, w.init=c(0.1,0.2, 0.7), lambda.init=c(5, 25, 40), epsilon=1e-5, max.iter=100)
ee = EM2
print(paste("Estimate pi = (", round(ee$w.curr[1],2), ",",
            round(ee$w.curr[2],2), ",",
            round(ee$w.curr[3],2), ")", sep=""))

print(paste("Estimate lambda = (", round(ee$lambda.curr[1],2), ",",
            round(ee$lambda.curr[2],2), ",",
            round(ee$lambda.curr[3],2), ")", sep=""))

plot(ee$log_liks, ylab='incomplete log-likelihood', xlab='iteration')
```

Check which estimators have the highest incomplete log-likelihood.
```{r}
EM1$log_liks[length(EM1$log_liks)]
EM2$log_liks[length(EM2$log_liks)]
```

Estimators from the two EM runs nearly have equally highest incomplete log-likelihoods. You can see that the incomplete log-likelihood is slightly higher for the estimators from the second EM run and so we will choose the estimators from the second EM run - $\hat\pi_1 = 0.25,\; \hat\pi_2 = 0.25,\; \hat\lambda_1 = 5.13,\; \hat\lambda_2 = 17.43,\; \hat\lambda_3 = 36.86$.
