---
title: "Assignment 1"
author: "James La Fontaine"
date: "26/08/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
```

## Question 1

### 1a)
```{r}
quiz = read.delim("quiz.txt", header = FALSE, sep = "")
responses = quiz[,1]
summary(responses)
sd(responses)
IQR(responses)
quantile(responses, type = 7)
boxplot(responses, horizontal = TRUE)
```

This distribution is asymmetrical and positively skewed with the centre lying roughly below the median 60. This distribution has a relatively large spread with a range of 233 due to an outlier, and even still has a range of approximately 140 when ignoring outliers. The distribution additionally has a relatively loose IQR of 45 and relatively large standard deviation of 40.54. The responses appear to spread out further and further as they become larger.

### 1b)

```{r}
# starting values for the parameters determined via the moment of methods
alpha.hat = mean(responses)^2/var(responses)
theta.hat = var(responses)/mean(responses)

gamma.fit = fitdistr(responses, densfun = "gamma", start = list(shape = alpha.hat, scale = theta.hat)) 

shape.hat = gamma.fit$estimate[[1]]
scale.hat = gamma.fit$estimate[[2]]
shape.hat
scale.hat
```

### 1c)

```{r}
hist(responses, freq = FALSE, col = "gray", main = NULL, xlab = "Response Values", nclass = 10, ylim = c(0,0.013))
curve(dgamma(x, shape = shape.hat, scale = scale.hat), lwd = 2, add = TRUE)
```

### 1d)

```{r}
gamma.sample = rgamma(116, shape = shape.hat, scale = scale.hat)


qqplot(gamma.sample, responses, main = "Gamma QQ Plot for Quiz Responses", xlab = "Theoretical Quantiles", ylab = "Sample Quantiles")

abline(a=0, b=1)
```

The model fits the data quite well and thus this QQ Plot demonstrates that the gamma distribution with the parameters estimated earlier is a good approximation for the distribution of the data.

## Question 4

Let Damjan's average of the sample minimum and maximum be Estimator 1. Let Julia's sample median be Estimator 2. Let Martina's sample mean be Estimator 3.

```{r}
numberofsimulations = 20000
N = numberofsimulations
estimator1 = 1:N
estimator2 = 1:N
estimator3 = 1:N
for (i in 1:N) {
  normal.sample = rnorm(10)
  estimator1[i] = (max(normal.sample) + min(normal.sample)) / 2
  estimator2[i] = median(normal.sample)
  estimator3[i] = mean(normal.sample)
}

# subtract 0 as this is the true mean of the standard normal distribution
bias.estimator1 = mean(estimator1) - 0
bias.estimator2 = mean(estimator2) - 0
bias.estimator3 = mean(estimator3) - 0

variance.estimator1 = var(estimator1)
variance.estimator2 = var(estimator2)
variance.estimator3 = var(estimator3)
```

```{r echo=FALSE}
cat("Bias of Estimator 1:", bias.estimator1)
cat("Bias of Estimator 2:", bias.estimator2)
cat("Bias of Estimator 3:", bias.estimator3)
cat("Variance of Estimator 1:", variance.estimator1)
cat("Variance of Estimator 2:", variance.estimator2)
cat("Variance of Estimator 3:", variance.estimator3)
boxplot(estimator1, estimator2, estimator3, names = c("Damjan (1)", "Julia (2)", "Martina (3)"), range = 0)


```

Note that the boxplot whiskers have been extended out to the maximums and minimums of each data set for improved visual clarity.