---
title: "Assignment 3"
author: "James La Fontaine"
date: "12/10/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

### 1ai)

H0: mX = mY

H1: mX ≠ mY
```{r question 1ai}
binom.test(12, 17)
```
The p-value > 0.05 so we cannot reject the null hypothesis that the median of X equals the median of Y at the significance level of 5%.

### 1aii)

H0: mX = mY

H1: mX ≠ mY

```{r question 1aii}
x = c(26.1, 26.6, 27.4, 27.5, 27.8, 28.1, 28.4, 29.5, 29.8, 30.4, 30.4, 31.2, 31.5, 32.9, 33.6, 34.1, 35.9)
y = c(27.4, 28.1, 22.9, 31.3, 16.3, 50.1, 20.0, 24.6, 23.3, 19.3, 24.4, 24.4, 29.5, 27.6, 21.7, 25.4, 39.4)

wilcox.test(x, y, paired = TRUE)


```
The p-value < 0.05 so we reject the null hypothesis at the median of X equals the median of Y at the significance level of 5% and can conclude that there is sufficient evidence to show that the location of X and Y differ.

### 1aiii)

H0: μX = μY

H1: μX ≠ μY

```{r question 1aiii}
x = c(26.1, 26.6, 27.4, 27.5, 27.8, 28.1, 28.4, 29.5, 29.8, 30.4, 30.4, 31.2, 31.5, 32.9, 33.6, 34.1, 35.9)
y = c(27.4, 28.1, 22.9, 31.3, 16.3, 50.1, 20.0, 24.6, 23.3, 19.3, 24.4, 24.4, 29.5, 27.6, 21.7, 25.4, 39.4)

t.test(x, y, paired = TRUE)
```
The p-value > 0.05 so we cannot reject the null hypothesis that the mean of X equals the mean of Y at the significance level of 5%.

## 1b)

The sign test has a larger type II error rate / smaller power and so it is plausible that the null hypothesis has incorrectly not been rejected in this test. For the t-test we have made the assumption that the differences between X and Y are normally distributed and since we have a small sample size of only 17, it is plausible that these differences do not follow a normal distribution. It would be more appropriate to give more consideration to the outcome of the Wilcoxon signed-rank test in this case, which simply assumes that the differences between X and Y are continuous and follow a symmetrical distribution, which is a reasonable assumption under the null hypothesis. Therefore, there is mild evidence that X and Y differ in location, however, further testing with a larger sample would be required to make stronger conclusions.

## 1c)

```{r question 1c}

B = 20000
n = 17
numRejectionsSign = 0
numRejectionsWilcoxon = 0
numRejectionsT = 0
  
for (i in 1:B) {
  numSuccesses = 0
  sampleDifference = rnorm(n, 3, 5)
  for (number in sampleDifference) {
    if (sign(number) == 1) {
      numSuccesses = numSuccesses + 1
    }
  }
  if (binom.test(numSuccesses, n)$p.value < 0.05) {
    numRejectionsSign = numRejectionsSign + 1
  }
  if (wilcox.test(sampleDifference)$p.value < 0.05) {
    numRejectionsWilcoxon = numRejectionsWilcoxon + 1
  }
  if (t.test(sampleDifference)$p.value < 0.05) {
    numRejectionsT = numRejectionsT + 1
  }
}

powerSign = numRejectionsSign / B
powerWilcoxon = numRejectionsWilcoxon / B
powerT = numRejectionsT / B

cat("Simulated power of sign test: ", powerSign, "\n")
cat("Simulated power of Wilcoxon test: ", powerWilcoxon, "\n")
cat("Simulated power of t-test: ", powerT, "\n")

```

## Question 2

### 2a)
```{r question 2a}
germinations = c(3, 4, 5, 6, 7, 8 ,9, 10, 11, 12, 13, 17)
count = c(1, 2, 2, 4, 10, 16, 9, 11, 13, 4, 7, 1)

experiments = data.frame(germinations, count)

data = rep(experiments$germinations, experiments$count)

p1 = prop.test(sum(data), 80*30)

p1

prop.estimate = as.numeric(p1$estimate)

```

### 2b)
```{r question 2b}
barplot(dbinom(3:17, 30, prop.estimate) * 80)
X1 <- cut(data, breaks = c(0, 5.5, 6.5, 7.5, 8.5, 9.5, 10.5, 11.5, 12.5, Inf))
T1 <- table(X1)
grouped.data <- as.numeric(T1)

p <- c(pbinom(5, 30, prop.estimate), dbinom(6:12, 30, prop.estimate), 1 - pbinom(12, 30, prop.estimate))

barplot(p * 80)

```

### 2c)
```{r question 2c}

chi1 = chisq.test(x=grouped.data, p = p)

chi1

X.squared = unname(chi1$statistic)

# recalculate the p-value using the correct degrees of freedom

1 - pchisq(unname(X.squared), length(grouped.data) - 2)

```

The p-value 0.5487 > 0.05 and so there is insufficient evidence to conclude that there is a difference between a Binomial distribution and the distribution of the number of germinations of seeds of the tested plant.

## Question 4

### 4a)
```{r question 4a}
Angle = c(rep(seq(0,30,10), each=10))

Panel = c(rep(rep(1:5, each = 2), 4))

Power = c(42.3, 41.4, 42.2, 40.3, 37.6, 35.7, 36.8, 34.9, 45.8, 43.7, 42.1, 40.2, 42.1,  40.3, 38.4, 36.5, 38.0, 37.1, 45.2, 43.1, 42.6, 40.8, 42.7, 40.8, 38.6, 36.7, 40.2, 38.3, 46.9, 44.8, 43.6, 41.5, 43.8, 41.9, 41.9, 39.8, 42.9, 40.8, 45.4, 43.5)

data = data.frame(Angle, Panel, Power)

model1 = lm(Power ~ factor(Angle) + factor(Panel), data = data)

anova(model1)

```

Let α = the effect of angle elevation on power output.

Let β = the effect of panel type on power output.

μij = μ + αi + βj

H0A: α1 = α2 = α3 = α4 = 0 

H1A: H ̅0

F value = 5.8632

p-value = 0.002614

Have assumed that:
- There is no statistical interaction between the factors and thus factor effects are additive.
- We have random samples drawn independently of each other from the different populations, each having a normal distribution
- All populations have the same variance, σ2

0.002614 < 0.05. Therefore, there is sufficient evidence to conclude at the 5% level of significance that the mean power output of solar panels varies between the different angles of elevation and thus that they have an effect on mean power output.

### 4b)
```{r question 4b}
model2 = lm(Power ~ factor(Angle) * factor(Panel), data = data)

anova(model2)

with(data, interaction.plot(Panel, Angle, Power, col = "blue"))
```
0.2045 > 0.05. Therefore, there is insufficient evidence to conclude at the the 5% level of significance that there is interaction between panel type and the angle of elevation.
